Intergenic regions

### 2016-11-15 AnalysisCode/intergenic_regions.Rmd
Move code for barplot

```{r}
genes <- import.gff("../RawData/TAIR10_GFF3_genes.gff.gz")
te <- import("../RawData/TAIR9_TE.bed.gz",
             format = "bed",
             extraCols = c(strand = "character",
                           TE = "character",
                           family = "character",
                           superfamily = "character"))
pseudogene <- import("../ProcessedData/GeneFeatures/pseudogene.bed.gz",
                     format = "bed",
                     extraCols = c(dot = "character",
                                   strand = "character",
                                   source = "character",
                                   type = "character",
                                   dot = "character",
                                   id = "character"))
dhs <- import("../RawData/Sullivan_DHS_PE_peaks_control.bed.gz",
              format = "bed",
              extraCols = c(dot = "character",
                            zero = "numeric",
                            second_dot = "character",
                            a = "numeric",
                            b = "numeric"))
centromeres <- import("../RawData/centromere_positions.txt",
                      format = "bed")

@@

```{r extend_coords}
# need to remove entries that cover the entire chromosome, they will overlap everything
chromosome_entries <- width(ranges(genes)) > 10^6

# check we have 5 chromosomes 
sum(chromosome_entries)

# now remove those entries
genes <- genes[!chromosome_entries]

# now extend 2 kb each direction
extended_genes <- genes + 2000
```

@@

Find the percentage of all intergenic regions that contain a TE insertion

```{r}
length(subsetByOverlaps(intergenic, tepid_coords)) / length(intergenic) * 100
```

Now try splitting up all intergenic regions longer than 1 kb. If longer than x kb, cut into x+1 equal bins, where x >= 1

```{r split_intergenic}
gr <- GRanges()
for(i in seq_along(intergenic)) {
        l <- width(intergenic[i])
        if(l < 1000) {
                gr <- c(gr, intergenic[i])
        } else {
                gr <- c(gr, unlist(tile(intergenic[i], n = floor(l / 1000))))
        }
}
```

```{r intersect_tiled_intergenic}
length(subsetByOverlaps(gr, tepid_coords)) / length(gr) * 100
```

Let's go with the original, non-cut, data.

>>>>

```{r}
genes <- import.gff("../RawData/TAIR10_GFF3_genes.gff.gz")
# filter so only type == gene
genes <- genes[genes$type == "gene"]
te <- import("../RawData/TAIR9_TE.bed.gz",
             format = "bed",
             extraCols = c(strand = "character",
                           TE = "character",
                           family = "character",
                           superfamily = "character"))
pseudogene <- import("../ProcessedData/GeneFeatures/pseudogene.bed.gz",
                     format = "bed",
                     extraCols = c(dot = "character",
                                   strand = "character",
                                   source = "character",
                                   type = "character",
                                   dot = "character",
                                   id = "character"))
dhs <- import("../RawData/Sullivan_DHS_PE_peaks_control.bed.gz",
              format = "bed",
              extraCols = c(dot = "character",
                            zero = "numeric",
                            second_dot = "character",
                            a = "numeric",
                            b = "numeric"))
centromeres <- import("../RawData/centromere_positions.txt",
                      format = "bed")

@@

```{r extend_coords}
extended_genes <- genes + 2000
```

@@

```{r}
fractionIntergenic <- length(subsetByOverlaps(intergenic, tepid_coords)) / length(intergenic) * 100
```

Write the intergenic regions to file

```{r write}
intergenic <- renameSeqlevels(intergenic, c("chr1", "chr2", "chr3", "chr4", "chr5"))
write_tsv(as_data_frame(intergenic), "../ProcessedData/GeneFeatures/intergenic_regions.bed", col_names = FALSE)
```

## Repeat all gene feature intersections

Need to repeat everything as some intervals were counted twice if two TEs were inserted in them. This is what we want for the insertion/deletion rare/common enrichments for fig 2C, D, but not for the barplot.

```{r}
fractionGene <- length(subsetByOverlaps(genes, tepid_coords)) / length(genes) * 100
fractionDHS <- length(subsetByOverlaps(dhs, tepid_coords)) / length(dhs) * 100
fractionTE <- length(subsetByOverlaps(te, tepid_coords)) / length(te) * 100
fractionPseudogene <- length(subsetByOverlaps(pseudogene, tepid_coords)) / length(pseudogene) * 100
```

Now need the other parts -- exon, intron, utr, upstream, downstream

```{r read_regions}
introns <- import("../ProcessedData/GeneFeatures/introns.bed.gz",
                  format = "bed")
exons <- import("../ProcessedData/GeneFeatures/exons.bed.gz",
                format = "bed")
utr3 <- import("../ProcessedData/GeneFeatures/utr3.bed.gz",
               format = "bed")
utr5 <- import("../ProcessedData/GeneFeatures/utr5.bed.gz",
               format = "bed")
upstream <- flank(genes, start = TRUE, both = FALSE, width = 2000)
downstream <- flank(genes, start = FALSE, both = FALSE, width = 2000)

# rename seq levels and drop metadata
introns <- dropSeqlevels(introns, c("chrM", "chrC"))
introns <- renameSeqlevels(introns, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
mcols(introns) <- NULL

exons <- dropSeqlevels(exons, c("chrM", "chrC"))
exons <- renameSeqlevels(exons, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
mcols(exons) <- NULL

utr3 <- renameSeqlevels(utr3, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
mcols(utr3) <- NULL

utr5 <- dropSeqlevels(utr5, c("chrM", "chrC"))
utr5 <- renameSeqlevels(utr5, c("Chr1", "Chr2", "Chr3", "Chr4", "Chr5"))
mcols(utr5) <- NULL
```

Get the intersections with TE variants

```{r}
fractionIntron <- length(subsetByOverlaps(introns, tepid_coords)) / length(introns) * 100
fractionExon <- length(subsetByOverlaps(exons, tepid_coords)) / length(exons) * 100
fractionUTR3 <- length(subsetByOverlaps(utr3, tepid_coords)) / length(utr3) * 100
fractionUTR5 <- length(subsetByOverlaps(utr5, tepid_coords)) / length(utr5) * 100
fractionUpstream <- length(subsetByOverlaps(upstream, tepid_coords)) / length(upstream) * 100
fractionDownstream <- length(subsetByOverlaps(downstream, tepid_coords)) / length(downstream) *100
```

```{r}
all_fractions <- c(fractionDHS, fractionUpstream, fractionUTR5, fractionExon, fractionIntron,
                   fractionUTR3, fractionDownstream, fractionPseudogene, fractionTE, fractionIntergenic)
barplot_names <- c("DHS", "Upstream", "5' UTR", "Exon", "Intron", "3' UTR", "Downstream", "Pseudogenes", "TE", "Intergenic")
intersection_df <- data.frame(all_fractions, barplot_names)
```

Barplot

```{r}
barplot(intersection_df$all_fractions, names.arg = intersection_df$barplot_names, las=2, ylim = c(0, 40),
        ylab = "% contining TE insertion", main = "Genomic features")
```

Save plot

```{r}
pdf("../Plots/barplot_genomic_feature_intersections.pdf", height = 4, width = 4)
barplot(intersection_df$all_fractions, names.arg = intersection_df$barplot_names, las=2, ylim = c(0, 40),
        ylab = "% contining TE insertion", main = "Genomic features")
dev.off()
```
